Phytoplankton Retention Mechanisms in Estuaries: A Case Study of the Elbe Estuary

### 2023-11-29 ms.tex
first review edit, to proof read for ross

\affil[1]{Universität Hamburg, Olbersweg 24, 22767 Hamburg, Germany}
\affil[2]{Cawthron Institute, 98 Halifax Street East,
Nelson 7010, New Zealand}

@@

Estuaries are complex and strongly dynamic systems such that it is still difficult to predict ecosystem dynamics or the effects of anthropogenic impacts due to their complex bathymetry \citep{Michael2016, Fringer2019}. 
% (Sensitivity analysis of three-dimensional salinity simulations in North San Francisco Bay using the unstructured-grid SUNTANS model,
% The future of coastal and estuarine modeling: Findings from a workshop)
Nevertheless, previous modelling studies have investigated potential retention mechanisms of phytoplankton in estuaries.
In \citet{Simons2006,Kimmerer2014} they used a Lagrangian model to study zooplankton retention.
\citet{Simons2006} examined the dispersal and flushing time of mussel larvae in the St. Lawrence Estuary while() \citep{Kimmerer2014} examined zooplankton movement in the San Francisco Estuary.
They were able to show that sinking and diel vertical migration slows the outwashing process and might be a beneficial retention strategy.
However, they did so by ignoring key processes like reproduction, mortality, and stranding or sedimentation processes.
Moreover, both studies were based on low resolution structured grid models, which we suspect to under-represent the complex bathymetry of estuarine systems \citep{Ye2018}.

@@

Here, we explore possible retention mechanisms of phytoplankton using the Elbe estuary as a case study.
It is located in the north of Germany and flows into the North Sea.
Like most alluvial estuaries, it is relatively shallow, with most of it averaging only a few meters in average depth.
Similar to other European estuaries it experienced a strong anthropogenic pressure over the last centuries.
Most notably diking to restrain it to a narrow channel and dredging to improve access to the Hamburg harbor.
Unlike other major European harbors, the Hamburg harbor is located far inwards roughly 100 \unit{km} away from the coast.
To create port access the main channel experience a sudden jump in bathymetry from approximately 5 \unit{m} at border of the city to up to 20 \unit{m} in the harbor and downstream (see fig. \ref{fig:bathymetry}). 
This bathymetric jump is suspected to be the cause of a collapse of the phytoplankton, resulting in an increase in oxygen depletion and high ammonium remineralization downstream of the bathymetric jump \citep{Schroeder1997,Holzwarth2018,Sanders2018}.
Ongoing dredging is being carried out to maintain the depth of the navigational channel causing high turbidity \citep{Kappenberg2001}.
While important aspects of the along-channel biochemical dynamics have been studied, little is known about their vertical and shore-to-shore dynamics \citep{Goosen1999, Dahnke2008, Sanders2018}.

\begin{figure}
    \includegraphics[width=8.3cm]{harbor_small.png}
    \caption{Bathymetry of the Elbe model around Hamburg. Note the bathymetric jumps from ~5 \unit{m} on the right, upstream side to a short 10 \unit{m} step in the upper harbor area to the ~20 \unit{m} in the lower harbor area all the way to the North Sea. Also note that there exists no channel that does not pass through the 10 \unit{km} of exclusively 20 \unit{m} deep channel.}
    \label{fig:bathymetry}
\end{figure}

@@

%\subsubsection{Lagrangian model}
In our study we use a Lagrangian approach with the particle tracking model OceanTracker \citep{Vennell2021}.
While off-line particle tracking on unstructured grids has been relatively computationally expensive until recently \citep{Vennell2021}, it offers several advantages.
Firstly, it allows us to reuse computationally expensive hydrodynamic models to model tracer-like objects.
This is overall much faster than recalculating the advection-diffusion-equation in an Eularian model.
Secondly, because we are simulating individually particles we are able to observe their tracks.
This makes the interpretation of our results more intuitive and allows us to include individual based properties and processes that can not or only indirectly be represented in Eulerian models.

%\subsubsection{Hydrodynamics data}
We use the hydrodynamics data of the most recent SCHISM model of the Elbe \citep{Pein2021}.
This model uses a three-dimensional unstructured  grid to represent the full Elbe estuary from the weir at Geesthacht to the North Sea, including several side-channels and the harbor area (see fig. \ref{fig:model_domain}). 
% (Look for more details in pein paper that might be interesting)
The model provides us with a node-based mesh containing a range of information such as water velocity, salinity, water level and dispersion.
The year represented in that dataset is 2012 with a temporal resolution of 1 hour and a dynamically varying spacial resolution with distance between nodes ranging from 5 to 1400 \unit{m} with a median distance of approximately 75 \unit{m}.

\begin{figure}
    \includegraphics[width=12cm]{model_domain_wcopyright.png}
    \caption{Map of the full model domain, with Geesthacht being the upstream boarder on the right and the North-sea being the downstream border on the left. The black outline marks the edge of the model domain. Blue and green dots show an example snapshot of a fraction of the phytoplankton in the model. The location of the initial release is shown in red. Blue represents floating, green particles stranded by the receding tide. The red area is the initial release location. The background map has been provided by OpenStreetMap.}
    \label{fig:model_domain}
\end{figure}

@@

Reproduction is represented as a chance of creating copies of themselves.
This is a novel feature applied in OceanTracker that has not been included in any previous Lagrangian model of this kind.
OceanTrackers recent advances in computational efficiency \citep{Vennell2021} and buffer handling make it possible to simulate a large number of particles over a long period of time for the first time on unstructured grids.
We perform several simulations for a range of reproduction rates that are constant over the lifetime of the particle and the chance to reproduce is evaluated every minute.
While a fixed reproduction rate is a simplification that does not allow for more realistic simulation of the population dynamics of a particular species, it does allow us to investigate the general mechanisms that enable plankton retention.

Mortality is induced either by high salinity, when they dry-out while stranded, or due to long term light limitation.
When particles are exposed to high salinity water above $20$PSU, a mortality probability of $0.5\%$ per minute is imposed.
This threshold is chosen based on a range of the salinity tolerances of estuarine phytoplankton species presented in \citep{vonAlvensleben2016}.
This is only an approximation and salinity tolerances many estuarine phytoplankton species deviate from this.
However, the main motivation for this choice is that most of the particles that die through this process have passed the isohaline for more than 12 hours, one tidal cycle, and are assumed not to return again through this isohaline.
Anything outside the 20 PSU isohaline is not considered part of the estuary for the purposes of this study.
Therefore, we are not tailoring our salinity tolerance to a specific species, but rather testing whether they can retain themselves within this isohaline.
We consider particles that are stranded outside the water by the receding tide, and lie dry for more than $7$ consecutive days to be dead and remove them.
Particles also die if they are light-limited for $28$ days \citep{Walter2017}.
They are considered light-limited at a depth of $1 \unit{m}$ based on turbidity data presented in \citep{Pein2021}.
A particle starts its life with a light budget of $28$ days, and each minute below $1 \unit{m}$ reduces this budget by one minute, while the opposite applies when they are above $1 \unit{m}$.
Children of light-limited parents inherit the remaining light budget of their parents.

@@

We investigate the effect of different patterns of vertical motion.
The first is monodirectional upward or downward vertical motion, representing either positively or negatively buoyant phytoplankton.
This buoyancy can be interpreted either as an active choice of buoyancy by the organism through adaptation, or as governed by the suspended matter aggregate on which they live.
For monodirectional vertical motion, we assign each particle a vertical velocity, which remains constant throughout its lifetime.
The second mode of vertical motion is diel vertical migration.
Here the particles change their direction of motion based on the current phase of the sun, creating a motion pattern where they rise during the day and sink during the night within the same velocity range.

% A large part of the Estuary is flat and regularly falls dry, making stranding and resuspension process important to predict their trajectories. 
We include a settling and resuspension model to represent tidal stranding and particles settling on the bed of the estuary.
Particles become stranded when the current grid cell becomes dry. 
They are not allowed to move from wet cells to dry cells, by the random walk dispersion applied to all particles.
A grid cell is considered dry based on the flag given in the SCHISM  hydrodynamic model output.
Once this cell is rewetted all stranded particles resuspend and are able to move again.
Particles settle on the bed once they attempt to move below the bottom model boundary and are resuspended based on a critical sheer velocity of 0.009 \unit{m s^{-1}} (see appendix for details).
The velocity profile in the bottom layer, or log layer, is calculated by
\begin{equation}
    U(z) = \frac{u_{*}}{\kappa} ln \frac{z}{z_{0}},
\end{equation}
where U is the friction velocity representing the drag at height z  above the seabed, $\kappa$ is the van Karman constant, $z_{0}$ is a length
scale reflecting the bottom roughness, and $u^{*}$ is the critical friction velocity.
If the friction velocity is above the critical friction velocity the particle is resuspended.
Particles that are stranded or settled on the bed are allowed to reproduce.
Particles are not only advected but also dispersed based on eddy diffusivity.
This allows us implement a dynamic dispersion that is crucial to represent tidal-pumping processes.
Dispersion was modeled using a random walk using a random number generator with a normal distribution.
Horizontally the standard distribution of the random walk was set to 0.1 \unit{m s^{-1}}.
The displacement by vertical dispersion $\partial z$ of particle $i$ is calculated by
\begin{equation}
    \partial z_{i} = K_{v}^{'}(z_{i}(n))\partial t + N(0, 2K_{v}(z_{i}))
\end{equation}
based on \citet{Yamazaki2014} where $z_{i}$ is the vertical position of the particle, $K_{v}^{'}$ is the vertical eddy diffusivity gradient, $K_{v}$ is the vertical eddy diffusivity and $N$ is the normal distribution.
The term based $K_{v}^{'}$ is needed to avoid particle accumulation on the top and bottom of the water column from the hydrodynamic model output.

For each particle we log their distance traveled, age, water depth, and status (whether they are drifting or settled on the river bank or bottom). This allows us for example to compare successfully retained particles (older than three months) unsuccessfully retained particles (dead after less than three months).
These observables are recorded every 12 hours starting at midnight.

%\subsubsection{Numerics}
Model simulations and visualizations were performed in Python making heavy use of Numba, a LLVM-based Python JIT compiler \citep{Lam2015} to significantly speed up the simulations \citep{Vennell2021}.
Trajectories were calculated using a second order Runge-Kutta scheme with a fixed time step of 60 \unit{s}.
Flow velocities, like any other hydrodynamic data, were interpolated linearly in time and space using barycentric coordinates, with the exception of water velocity in the bottom model cell, where logarithmic vertical interpolation is used.

@@

In the first experiment, we examine a range of different monodirectional upward or downward particle velocities from $-10$ to $+10$ \unit{mm\;s^{-1}} in $2$ \unit{mm\;s^{-1}} steps representing sinking or rising phytoplankton organisms \citep{Fennessy1996}.
Each vertical velocity is examined for a range of different reproduction rates, expressed as population doubling times in idealized conditions ranging from 40 to 404 days with a logarithmic scaling.
In the following, we will use reproduction rate to refer to the prescribed population growth rate under ideal conditions and use growth rate whenever we describe population growth in nature. 
Hence, a total of 187 different scenarios are tested.
In the second set of model experiments, we study the influence of possible diel vertical migration patterns for the same vertical velocities and reproduction rates.

In both sets of experiments, we release 10,000 individuals representing the studied phytoplankton population at the beginning of the year.
This results in over 1 billion individual particles simulated for each case with approximately 1 million particles active simultaneously counted over all cases over 500,000 time steps.
The initial population is homogeneously distributed in a volume covering the full water column at the weir in Geesthacht (see fig. \ref{fig:model_domain}) and examine how the population distributes itself over the estuary and whether it is able to maintain its population size over time.
Conceptually, we consider a population to be successfully retained if it is showing long term growth. 
Practically, this is evaluated by comparing the population size at the end of the year to the size after release.
The first three months of the simulations are considered as initial model spin-up time during which the initial population is dispersed downstream throughout the estuary.
Population size changes are measured at the end of the year relative to the population size after this initial spin-up time.

@@

Our simulations show that the population is able to successfully retains itself under certain conditions.
Passively drifting particles are able to sustain themselves in the estuary if they have a reproduction rate that doubles their population size within approximately 3 months.
Note that the growth rates realized in nature may vary from this value due to e.g. nutrient or temperature limitations. The reproduction thresholds should be interpreted as an upper bound rather than an accurate estimate of the growth rate.

For the case of the monodirectional movement we see that a higher positive velocity (representing buoyancy) and higher reproduction rates are more beneficial for retention success than a downward oriented velocity (sinking) and lower reproduction rates.
As expected, simulations in which the reproduction is set to zero do not show any retention success.
While it is easy to understand that high reproduction rates aid retention, we were surprised that buoyant phytoplankton particles are more successful in maintaining their growth in an estuary than sinking ones.

@@

We are now taking a closer look at spacial factors that allow phytoplankton particles to maintain net growth in the estuary. 
For this analysis we used data from both sets of experiments.
Fig. \ref{fig:migration-long-vs-short} compares two box plots showing the average water depth at the location of each particle between particles that remained alive for less than three months (short-living) and for more than three months (long-living).
Depth is measured relative to the current water surface.
Hence, a value above zero indicate that the particle is stranded on the river bank.
These analyses demonstrate that long-living particles predominantly live close to the river banks in shallower waters or on tidal flats.

\begin{figure}
    \includegraphics[width=8.3cm]{retention_boxplot.png}
    \caption[]{Box and violin plot showing the vertical distribution of particles that are passively drifting.  Short-living are those younger than 3 months and long-living all those older than that. Depth is measured relative to the current water surface with positive numbers being above the water surface i.e. stranded on shore.}
    \label{fig:migration-long-vs-short}
\end{figure}

We moreover analyzed the horizontal spacial distribution of long and short-living particles in fig. \ref{fig:migration-long-vs-short-heatmap}.
To do this, we divide the model domain into equally sized hexagons.
The color of each hexagon indicates the average age of the particles within it.
Hexagons with a yellow color indicate an average age of over three months.
These yellow areas are mainly found along the river banks in shallow waters or tidal flats.

\begin{figure}
    \includegraphics[width=12cm]{age_hexbin.png}
    \caption[]{
        Hex-bin heatmap of the Elbe with Hamburg in the bottom right showing average particle age per bin.
        Colors indicate the age of the particles, with yellowish colors indicating an average age of over three months.
        Yellow areas are mainly found along the river banks in shallow waters or tidal flats.
        The important areas are Mühlenberger Loch (a), Wedeler Marsch (b) Haseldorfer Binnenelbe (c), Asseler- and Schwarztonnensand (d), at the mouth of Wischhafener Süderelbe (f), , and Stör (e), and at Nordkedding (g) and Neufelder Marsch (h).
    }
    \label{fig:migration-long-vs-short-heatmap}
\end{figure}

To further investigate the reasons for the positive effect of buoyancy and the importance of shallow waters and tidal flats, we repeated the first set of simulations and disabled the reproduction of settled and stranded particles.
Under these conditions, populations were unable to retain themselves in the estuary, regardless of their vertical velocity and reproduction rate indicating that tidal flats are essential for the survival of the population.

@@

While our model does have a settling and resuspension mechanic based on critical sheer velocities we still assume a static bathymetry with sediments not able to move or bury particles. 
This masks potential losses due to particles being buried but also decreases resuspension times.

@@

% \appendix

% \section{}    %% Appendix A

@@

% \appendixfigures  %% needs to be added in front of appendix figures

\appendixtables   %% needs to be added in front of appendix tables

>>>>

\affil[1]{Institute of Marine Ecosystem and Fishery Science, Universität Hamburg, Olbersweg 24, 22767 Hamburg, Germany}
\affil[2]{Cawthron Institute, 98 Halifax Street East,
Nelson 7010, New Zealand}

@@

Estuaries are complex and strongly dynamic systems such that it is still difficult to predict ecosystem dynamics or the effects of anthropogenic impacts due to their complex bathymetry \citep{Michael2016, Fringer2019}. 
% (Sensitivity analysis of three-dimensional salinity simulations in North San Francisco Bay using the unstructured-grid SUNTANS model,
% The future of coastal and estuarine modeling: Findings from a workshop)
Nevertheless, there are sophisticated estuarine models that are able to reproduce the complex dynamics of estuaries reasonably well. This includes currents and water levels on the physical side, but also chlorophyll concentrations and other biologically driven properties \citep{Pein2021, Schol2014}.
However, these are Eulerian models.
This means that they are based on a fixed grid and calculate the concentration of a tracer, such as phytoplankton, at each grid cell.
This makes it difficult to study concepts such as retention times, as they lack temporal consistency, meaning that the life history and trajectory of a phytoplankton cell cannot be tracked.
Previous modeling studies have attempted to overcome this problem using a Lagrangian approach.
A Lagrangian model does not try to track e.g. concentrations at fixed positions, but rather follows the motion of individual particles that can be used to represent e.g. water parcels or organisms.
In \citet{Simons2006,Kimmerer2014} they used a Lagrangian model to study zooplankton retention.
\citet{Simons2006} examined the dispersal and flushing time of mussel larvae in the St. Lawrence Estuary while() \citep{Kimmerer2014} examined zooplankton movement in the San Francisco Estuary.
They were able to show that sinking and diel vertical migration slows the outwashing process and might be a beneficial retention strategy.
However, they did so by ignoring key processes like reproduction, mortality, and stranding or sedimentation processes.
Moreover, both studies were based on low resolution structured grid models, which we suspect to under-represent the complex bathymetry of estuarine systems \citep{Ye2018}.

@@

Here, we explore possible retention mechanisms of phytoplankton using the Elbe estuary as a case study.
It is located in the north of Germany and flows into the North Sea.
Like most alluvial estuaries, it is relatively shallow, with most of it averaging only a few meters in average depth.
Similar to other European estuaries it experienced a strong anthropogenic pressure over the last centuries.
Most notably diking to restrain it to a narrow channel and dredging to improve access to the Hamburg harbor.
Unlike other major European ports, the port of Hamburg is located far inwards roughly 100 \unit{km} behind the coastline.
To create port access the main channel is dredged and experiences a sudden jump in bathymetry from approximately 5 \unit{m} at border of the city to up to 20 \unit{m} in the port and downstream (see fig. \ref{fig:bathymetry}). 
This bathymetric jump is suspected to be the cause of a collapse of the phytoplankton, resulting in an increase in oxygen depletion and high ammonium remineralization downstream of the bathymetric jump \citep{Schroeder1997,Holzwarth2018,Sanders2018}.
Ongoing dredging is being carried out to maintain the depth of the navigational channel causing high turbidity \citep{Kappenberg2001}.
While important aspects of the along-channel biochemical dynamics have been studied, little is known about their vertical and shore-to-shore dynamics \citep{Goosen1999, Dahnke2008, Sanders2018}.

\begin{figure}
    \includegraphics[width=8.3cm]{harbor_small.png}
    \caption{Bathymetry of the Elbe model around Hamburg. Note the bathymetric jumps from ~5 \unit{m} on the right, upstream side to a short 10 \unit{m} step in the upper port area to the ~20 \unit{m} in the lower port area all the way to the North Sea. Also note that there exists no channel that does not pass through the 10 \unit{km} of exclusively 20 \unit{m} deep channel.}
    \label{fig:bathymetry}
\end{figure}

@@

%\subsubsection{Lagrangian model}
In our study we use a Lagrangian approach with the particle tracking model OceanTracker \citep{Vennell2021}.
While off-line particle tracking on unstructured grids has been relatively computationally expensive until recently \citep{Vennell2021}, it offers several advantages.
Firstly, it allows us to reuse computationally expensive hydrodynamic models to model tracer-like objects.
This is overall much faster than recalculating the advection-diffusion-equation in an Eularian model.
Secondly, because we are simulating individually particles we are able to observe their tracks.
In our model, we use these particles to represent phytoplankton cells or aggregates.
The temporal consistency of a Lagrangian model, the fact that we know the history of each particle, makes the interpretation of our results more intuitive and allows us to include individual-based properties and processes that cannot be represented in Eulerian models like e.g. retention times.

%\subsubsection{Hydrodynamics data}
We use the hydrodynamic data generated by the latest SCHISM model of the Elbe estuary \citep{Pein2021} from the weir at Geesthacht to the North Sea, including several side channels and the port area (see figure \ref{fig:model_domain}).
SCHISM solves the Reynolds-averaged Navier-Stokes equations on unstructured meshes assuming hydrostatic conditions with a time step of 60 \unit{s}.
The unstructured mesh is three-dimensional and consists of 32k nodes using terrain-following coordinates based on the LSC2 technique \citep{Zhang2016} for the vertical grid, allowing a maximum number of 20 levels.
Regions with depths less than 2 \unit{m} are resolved by only one vertical level.
The bathymetric data were provided by the German Federal Maritime and Hydrographic Agency (Bundesamt fuer Seeschifffahrt und Hydrographie, BSH) and the German Waterways Agency (Wasserstraßen- und Schiffahrtsamt, WSA) with a horizontal resolution of 50 m in the German Bight, 10 m in the Elbe estuary and 5 m in the Hamburg port \cite{Stanev2019}.
The boundary conditions on the seaward side include sea surface elevation, horizontal currents, salinity and temperature \cite{Stanev2019} and discharge and temperature from the Elbe river on the land-ward side.
Atmospheric forcing includes wind, air temperature, precipitation, shortwave and longwave radiation \cite{Stanev2019}.
Model validation is based on tide gauge stations and long-term stationary measurements of salinity, water temperature, and horizontal currents.
Biochemical variables, including chlorophyll, are based on long-term measurements at the Seemannshöft and Grauerort stations \cite{Pein2021}.
The model provides us with a node-based mesh containing a range of information such as water velocity, salinity, water level and dispersion.
The year represented in that dataset is 2012 with a temporal resolution of 1 hour and a dynamically varying spacial resolution with distance between nodes ranging from 5 to 1400 \unit{m} with a median distance of approximately 75 \unit{m}.

\begin{figure}
    \includegraphics[width=12cm]{model_domain_wcopyright.png}
    \caption{Map of the full model domain, with Geesthacht being the upstream boarder on the right and the North-sea being the downstream border on the left. The black outline marks the edge of the model domain. Blue and green dots show an example snapshot of a fraction of the phytoplankton in the model. The location of the initial release is shown in red. Blue represents floating, green phytoplankton stranded by the receding tide. The red area is the initial release location. The background map has been provided by © OpenStreetMap contributors 2023. Distributed under the Open Data Commons Open Database License (ODbL) v1.0.}
    \label{fig:model_domain}
\end{figure}

@@

Reproduction is represented as a fission process, where each phytoplankton cell has a probability to split effectively producing a copy.
This is a novel feature applied in OceanTracker that has not been included in any previous Lagrangian model of this type.
OceanTracker's recent advances in computational efficiency \citep{Vennell2021} and buffer handling make it possible for the first time to simulate a large number of particles over a long period of time on unstructured grids.
We perform multiple simulations for a range of reproduction rates, implemented as a fission probability evaluated every minute, that are constant over the lifetime of the cell.
While a fixed reproduction rate is a simplification that does not allow for more realistic simulation of the population dynamics of a particular species, it does allow us to investigate the general mechanisms that enable plankton retention.

Mortality is induced by one of three processes: high salinity, when they dry out while stranded, or due to long term light limitation.
When particles cells are exposed to high salinity water above $20$PSU, a mortality probability of $0.5\%$ per minute is applied removing dead phytoplankton cells from the simulation (see salinity map in fig. \ref{fig:salinity} ).
This threshold is chosen based on a range of the salinity tolerances of estuarine phytoplankton species presented in \citep{vonAlvensleben2016}.
This is only an approximation and salinity tolerances many estuarine phytoplankton species deviate from this.
However, the main motivation for this choice is that most of the phytoplankton cells that die through this process have passed the isohaline for more than 12 hours, one tidal cycle, and are assumed not to return again through this isohaline.
Anything outside the 20 PSU isohaline is not considered part of the estuary for the purposes of this study.
Therefore, we are not tailoring our salinity tolerance to a specific species, but rather testing whether they can retain themselves within this isohaline.
We consider phytoplankton cells that are stranded out of the water by the receding tide, and lie dry for more than $7$ consecutive days to be dead and remove them.
Note that these dry cells are not necessarily completely devoid of water, but are considered dry if the majority of its area has a water level is of below 0.1 \unit{m}.
Additionally, in nature these areas typically contain small sub-resolution structures like tidal ripples or small puddles and often also vegetation.
There are currently no studies investigating the time range for survival of stranded phytoplankton on tidal-flats or marshes in estuaries.
Therefore, we performed a sensitivity analysis to determine the effect of this parameter on the retention success of the phytoplankton population (see section \ref{sec:dry-out}).
Phytoplankton cells will also die if they are light-limited for $14$ days.
This value is based on measurements presented in \citep{Walter2017} which imply that the majority of phytoplankton is dead after $14$ days of light limitation.
A sensitivity analysis for this parameter is presented in sec. \ref{sec:light-limitation} suggesting no strong influence on the retention success.
They are considered light-limited below a depth of $1 \unit{m}$ based on SPM data presented in \citep{Stanev2019}.
The initial batch of phytoplankton cells starts their life with a full light budget of $14$ days, and each minute below $1 \unit{m}$ reduces this budget by one minute, while the opposite applies if they are above $1 \unit{m}$.
When a cell splits both inherit the same remaining light budget.

@@

We investigate the effect of different patterns of vertical motion.
The first is monodirectional upward or downward vertical motion, representing either positively or negatively buoyant phytoplankton.
This buoyancy can be interpreted either as an active choice of buoyancy by the organism through adaptation, or as governed by the suspended matter aggregate on which they live.
For monodirectional vertical motion, we assign each phytoplankton cell a vertical velocity, which remains constant throughout its lifetime.
The second mode of vertical motion is diel vertical migration.
Here phytoplankton cells change their direction of motion based on the current phase of the sun, creating a motion pattern where they rise during the day and sink during the night.
This behavior is often assumed to be performed to maximize light capture while avoiding predation - or as we suspect increase retention.

% A large part of the Estuary is flat and regularly falls dry, making stranding and resuspension process important to predict their trajectories. 
We include a settling and resuspension model to represent tidal stranding and phytoplankton cells settling on the bed of the estuary.
Stranding phytoplankton and microphytobenthos have been shown on several occasions to be a major driver of estuarine primary production \citep{Carlson1984,DeJonge1992,Kromkamp1995,Savelli2019}.
Phytoplankton cells become stranded when the current grid cell becomes dry and stay in place until they are resuspended or dry-out. 
They are not allowed to move from wet cells to dry cells, by the random walk diffusion applied to all phytoplankton cells.
A grid cell is considered dry based on the flag given in the SCHISM  hydrodynamic model output.
Once this cell is becomses flooded again all stranded phytoplankton cells resuspend and are able to move again.
Phytoplankton cells settle on the bed once they attempt to move below the bottom model boundary and are resuspended based on a critical sheer velocity of 0.009 \unit{m s^{-1}}.
The velocity profile in the bottom layer, or log layer, is calculated by
\begin{equation}
    U(z) = \frac{u_{*}}{\kappa} ln \frac{z}{z_{0}},
\end{equation}
where U is the friction velocity representing the drag at height z  above the seabed, $\kappa$ is the van Karman constant, $z_{0}$ is a length
scale reflecting the bottom roughness, and $u^{*}$ is the critical friction velocity.
If the friction velocity is above the critical friction velocity the phytoplankton cell is resuspended.
Phytoplankton cells that are stranded or settled on the bed are allowed to reproduce.
Phytoplankton cells are not only advected but also diffused based on eddy diffusivity which is crucial to represent tidal-pumping processes.
Diffusion was modeled using a random walk using a random number generator with a normal distribution.
Horizontally the standard distribution of the random walk is set to 0.1 \unit{m s^{-1}}.
The vertical displacement of a phytoplankton cell $\partial z$  $i$ is calculated by
\begin{equation}
    \partial z_{i} = K_{v}^{'}(z_{i}(n))\partial t + N(0, 2K_{v}(z_{i}))
\end{equation}
based on \citet{Yamazaki2014} where $z_{i}$ is the vertical position of the phytoplankton cell, $K_{v}^{'}$ is the vertical eddy diffusivity gradient, $K_{v}$ is the vertical eddy diffusivity and $N$ is the normal distribution.
The term based $K_{v}^{'}$ is needed to avoid phytoplankton accumulation on the top and bottom of the water column from the hydrodynamic model output.

For each phytoplankton cell we log their distance traveled, age, water depth, and status (whether they are drifting or settled on the river bank or bottom). This allows us for example to compare successfully retained phytoplankton cell (older than three months) unsuccessfully retained phytoplankton cell (dead after less than three months).
These observables are recorded every 12 hours starting at midnight.

%\subsubsection{Numerics}
Model simulations and visualizations were performed in Python making heavy use of Numba, a LLVM-based Python JIT compiler \citep{Lam2015} to significantly speed up the simulations \citep{Vennell2021}.
Trajectories were calculated using a second order Runge-Kutta scheme with a fixed time step of 60 \unit{s}.
Flow velocities, like any other hydrodynamic data, were interpolated linearly in time, linearly in space on the vertical axis and on the horizontal axis using barycentric coordinates, with the exception of water velocity in the bottom model cell, where logarithmic vertical interpolation is used.

@@

In the first experiment, we examine a range of different monodirectional upward or downward particle velocities from $-10$ to $+10$ \unit{mm\;s^{-1}} in $2$ \unit{mm\;s^{-1}} steps representing sinking or rising phytoplankton organisms \citep{Fennessy1996}.
Each vertical velocity is examined for a range of different reproduction rates, expressed as population doubling times in ranging from 40 to 404 days with a logarithmic scaling.
In the following, we will use reproduction rate to refer to the prescribed population growth rate under ideal conditions and use growth rate whenever we describe population growth in nature. 
Hence, a total of 187 different scenarios are tested.
In the second set of model experiments, we study the influence of possible diel vertical migration patterns for the same vertical velocities and reproduction rates.

In both sets of experiments, we release 10,000 individuals representing the studied phytoplankton population at the beginning of the year.
This results in over 1 billion individual particles simulated for each case with approximately 1 million particles active simultaneously counted over all cases over 500,000 time steps.
This corresponds approximately to a one to one ratio of simulated phytoplankton cells to mesh nodes in the hydrodynamic model at each time step.
The initial population is homogeneously distributed in a volume covering the full water column at the weir in Geesthacht (see fig. \ref{fig:model_domain}) and examine how the population distributes itself over the estuary and whether it is able to maintain its population size over time.
Conceptually, we consider a population to be successfully retained if it is able to sustain itself long term or even shows growth.
Practically, this is evaluated by comparing the population size at the end of the year to the size after release.
The choice of one year is considered reasonable because it covers the full seasonal cycle and is also much longer than the average exit or flushing time of the estuary (see \ref{fig:flushing-heatmap}).
The first three months of the simulations are considered as initial model spin-up time during which the initial population is dispersed downstream throughout the estuary.
Population size changes are measured at the end of the year relative to the population size after this initial spin-up time.

@@

Our simulations show that the population is able to successfully retains itself under certain conditions.
Passively drifting phytoplankton is able to sustain themselves in the estuary if they have a reproduction rate that doubles their population size within approximately 3 months.
Note that the growth rates realized in nature may vary from this value due to e.g. nutrient or temperature limitations. The reproduction thresholds should be interpreted as an upper bound rather than an accurate estimate of the growth rate.

For the case of the monodirectional movement we see that a higher positive velocity (representing buoyancy) and higher reproduction rates are more beneficial for retention success than a downward oriented velocity (sinking) and lower reproduction rates.
As expected, simulations in which the reproduction is set to zero do not show any retention success.
While it is easy to understand that high reproduction rates aid retention, we were surprised that buoyant phytoplankton cells are more successful in maintaining their growth in an estuary than sinking ones.

@@

We are now taking a closer look at spacial factors that allow phytoplankton cells to maintain net growth in the estuary. 
For this analysis we used data from both sets of experiments.
Fig. \ref{fig:migration-long-vs-short} compares two box plots showing the average water depth at the location of each phytoplankton cell between those cells that remained alive for less than three months (short-living) and for more than three months (long-living).
Depth is measured relative to the current water surface.
Therefore, a value greater than zero indicates that the phytoplankton cell is stranded on the shore during ebb tide.
For reference, the water level varies on average by about 5 \unit{m} due to the tides. \citep{Stanev2019, Schol2014}.
These analyses show that long-living phytoplankton predominantly live close to the river banks in shallower waters or on tidal flats.

\begin{figure}
    \includegraphics[width=8.3cm]{retention_boxplot.png}
    \caption[]{Box and violin plot showing the vertical distribution of phytoplankton that are passively drifting.  Short-living are those younger than 3 months and long-living all those older than that. Depth is measured relative to the current water surface with positive numbers being above the water surface i.e. stranded on shore.}
    \label{fig:migration-long-vs-short}
\end{figure}

We moreover analyzed the horizontal spacial distribution of long and short-living phytoplankton in fig. \ref{fig:migration-long-vs-short-heatmap}.
To do this, we divide the model domain into equally sized hexagons.
The color of each hexagon indicates the average age of the phytoplankton cells within it calculated across all cases.
Note, that the spatial age structure is similar for all cases.
Hexagons with a yellow color indicate an average age of over three months.
These yellow areas are mainly found along the river banks in shallow waters or tidal flats.

\begin{figure}
    \includegraphics[width=12cm]{age_hexbin.png}
    \caption[]{
        Hex-bin heatmap of the average age of phytoplankton cells in the Elbe estuary across all cases.
        The Hamburgs port area is located in the bottom right with the North Sea to the left.
        Colors indicate the age of the phytoplankton, with yellowish colors indicating an average age of over three months.
        Yellow areas are mainly found along the river banks in shallow waters or tidal flats.
        The important areas are Mühlenberger Loch (a), Wedeler Marsch (b) Haseldorfer Binnenelbe (c), Asseler- and Schwarztonnensand (d), at the mouth of Wischhafener Süderelbe (f), , and Stör (e), and at Nordkedding (g) and Neufelder Marsch (h).
    }
    \label{fig:migration-long-vs-short-heatmap}
\end{figure}

For comparison, we show the average exit time for each hexagon until the 20 \unit{PSU} isohaline is reached in \ref{fig:flushing-heatmap}.
This calculation is based on a separate simulation where we released approximately 1.8 million particles homogeneously distributed over the estuary.
We released one batch during winter in high discharge discharge conditions at the first of January and another batch during the low summer discharge at the first of July.
We disabled reproduction, light-limitation, stranding and settling on the riverbed for this simulation to isolate the effect of advection and dispersion.
\begin{figure}
    \includegraphics[width=12cm]{flush_hexbin.png}
    \caption[]{
        Hex-bin heatmap showing average exiting times of the Elbe estuary with Hamburgs port area, as shown in \ref{fig:bathymetry}, in the bottom right without reproduction, light-limitation, stranding and settling on the riverbed.
        Colors indicate the time of a phytoplankton cell or water parcel to reach the 20 \unit{PSU} isohaline from its origin hexagon.
    }
    \label{fig:flushing-heatmap}
\end{figure}

To further investigate the reasons for the positive effect of buoyancy and the importance of shallow waters and tidal flats, we repeated the first set of simulations and disabled the reproduction of settled and stranded phytoplankton.
Under these conditions, populations were unable to retain themselves in the estuary, regardless of their vertical velocity and reproduction rate indicating that tidal flats are essential for the survival of the population.

@@

While our model does have a settling and resuspension mechanic based on critical sheer velocities we still assume a static bathymetry with sediments not able to move or bury phytoplankton. 
This masks potential losses due to phytoplankton being buried but also decreases resuspension times.

@@

\appendix

\section{Sensitivity analysis "dry-out"}    %% Appendix A
\label{sec:dry-out}

% v26 1 day
% v27 14 days

In figure \ref{fig:retention_success_short_stranding} and \ref{fig:retention_success_long_stranding} we present the results for a sensitivity analysis of the mortality threshold due to stranding i.e. drying out of 1 day and 14 days compared to the 7 days presented in fig. \ref{fig:retention_success}

\section{Sensitivity analysis "light-limitation"}
\label{sec:light-limitation}

% v28 28 days dry out
% v29 7 days

In figure \ref{fig:retention_success_low_light} and \ref{fig:retention_success_high_light} we present the results for a sensitivity analysis of the mortality threshold due to light limitation of 7 day and 28 days compared to the 14 days presented in fig. \ref{fig:retention_success}.

\section{Salinity}
\label{sec:salinity}

In figure \ref{fig:salinity} shows a map of average salinity of the Elbe estuary. Salinity is averaged in depth and over the whole year.

@@

\appendixfigures  %% needs to be added in front of appendix figures
